---
title: "Big Data Bowl Data Cleaning and Manipulation"
author: "Kai Franke"
date: "2023-12-20"
output: html_document
---

This notebook contains the data cleaning processes for the Big Data Bowl 2024 for the UMN Sports Analytics Club.

# Data Cleaning and Manipulation

## Loading in Data

To load in the data, we simply just load the csv's. Then to get all of the tracking data, I just did a for loop through all of the csv's in the file. I also loaded in all of the necessary libraries.

```{r Loading Libraries and Data}

library(tidyverse)
library(dplyr)
library(stringr)
library(ggplot2)

setwd("C:/Users/kaifr/OneDrive/Miscellaneous/Desktop/Productivity/BDB2024/nfl-big-data-bowl-2024")
file_list = list.files(path = "C:/Users/kaifr/OneDrive/Miscellaneous/Desktop/Productivity/BDB2024/nfl-big-data-bowl-2024")

games = read.csv(file = "games.csv")
players = read.csv(file = "players.csv")
plays = read.csv(file = "plays.csv")
tackles = read.csv(file = "tackles.csv")

tracking = read.csv(file = "tracking_week_1.csv")

for (i in 6:14) {
  
  print(file_list[i])
  tracking = rbind(tracking, read.csv(file = file_list[i]))
  
}

```

## Plays Data Edit

These lines of code make it easier to understand what each of the pass results mean. I changed them from a simple letter to the entirety of what each letter means.

```{r}

plays = plays %>%
  mutate(passResult = ifelse(passResult == "C", "Complete pass", passResult),
         passResult = ifelse(passResult == "I", "Inomplete pass", passResult),
         passResult = ifelse(passResult == "S", "QB sack", passResult),
         passResult = ifelse(passResult == "IN", "Intercepted pass", passResult),
         passResult = ifelse(passResult == "R", "Scramble", passResult))

```


## Merging Data

To merge the data correctly, I needed to investigate each of the data frames. I looked into the column names and see how they correspond to each of the data sets. I also took a peek into the repository from Kaggle to see what each of the data sets have and what they mean.

For the tackling information, I added that when the information was NA, that it is a 0. This means that the player was not a part of the tackle. I also created a column called 'partOfTackle' to signify whether or not the player had anything to do with the tackle attempt.

I then removed the excess data sets.

```{r Merging}

colnames(players)
colnames(plays)
colnames(tackles)
colnames(tracking)

playerwTrack = left_join(tracking, players)
PWTtack = left_join(playerwTrack, tackles)

PWTtack = PWTtack %>%
  mutate(tackle = ifelse(is.na(tackle), 0, tackle),
         assist = ifelse(is.na(assist), 0, assist),
         forcedFumble = ifelse(is.na(forcedFumble), 0, forcedFumble),
         pff_missedTackle = ifelse(is.na(pff_missedTackle), 0, pff_missedTackle),
         partOfTackle = ifelse(tackle + assist + forcedFumble + pff_missedTackle > 0, "Y", "N"))

trackGame = right_join(games, PWTtack)
track = left_join(trackGame, plays)

rm(games, players, playerwTrack, playTack, PWTtack, tackles, trackGame, tracking, file_list, i)

```


## Standardizing

The next step was to standardize the data depending on where the play was going. This makes it easier for analysis later on.

```{r Quick Clean}

track2 = track %>%
  arrange(gameDate, gameId, playId, displayName, frameId) %>%
  distinct() %>%
  mutate(x = ifelse(playDirection == "left", 120 - x, x),
         y = ifelse(playDirection == "left", 160 / 3 - y, y),
         dir = ifelse(playDirection == "left", dir + 180, dir),
         dir = ifelse(dir > 360, dir - 360, dir),
         o = ifelse(playDirection == "left", o + 180, o),
         o = ifelse(o > 360, o - 360, o))



```

## Looking at a play

Next, I wanted to make sure the data looked right and looked at an example play to verify this and see if it made sense.

```{r}

set.seed(2010)
example_play = plays %>%
  dplyr::select(gameId, playId, playDescription) %>% 
  sample_n(1)
example_play

example_play_tracking_data = track2 %>%
  inner_join(example_play, by = c("gameId", "playId"))


ggplot(example_play_tracking_data, aes(x, y, color = partOfTackle)) +
  geom_point() +
  xlim(0,120) +
  ylim(0, 53.3) +
  coord_fixed()

```


```{r}

wide_data = track2 %>%
     pivot_wider(
         id_cols = c(frameId, gameId, playId),
         names_from = position,
         values_from = c(x, y),
         names_glue = "{.value}_{position}"
     )

```


